---
title: "DATASCI 306"
subtitle: "Lecture 13"
format: live-revealjs
engine: knitr
resources:
  - data/a2housing.RData
webr:
  packages:
    - tidyverse
    - sf
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(sf)
load("data/a2housing.RData")
theme_set(theme_classic())
options(tigris_use_cache = TRUE)
```
```{webr}
#| echo: false
#| message: false
#| warning: false
#| autorun: true
library(tidyverse)
library(sf)
load("data/a2housing.RData")
theme_set(theme_classic())
```

```{=html}
<style>
.qrcode img {
  margin: 20px auto;
}
</style>
```

[https://ds306.org/lectures/week8/lecture13.html](https://ds306.org/lectures/week8/lecture13.html)

# Where should I live?

- Soon you will graduate college.
- You may choose to live anywhere!
- How to decide where?

## Factors to consider

- Cost of living
- Job opportunities
- Climate
- Culture
- ???

## Survey

https://ds306.org/survey3

{{< qrcode https://ds306.org/survey3 width=300 height=300 >}}

## Spatial data analysis
- Spatial data analysis is the study of data that has a spatial component, such as location or geography.
- Spatial data analysis can help us understand patterns and relationships in data that are influenced by location.

## Example spatial data set: housing prices in A2
- Suppose you choose to remain in Ann Arbor. 
- You may want to buy a house.
- Where should you buy a house?

## `a2housing`
`a2housing` contains information about housing prices in Ann Arbor over the past
4 years.

```{webr}
a2housing
```

## Basics
- `a2housing` is a data frame with ~11k rows and 10 columns.
- Each row is a house that was sold in Ann Arbor between 2021 and present.

## Exploring the data
```{r echo=T, messages=F}
library(GGally)
a2housing |> select(where(is.numeric)) |> ggpairs()
```

## Correlations
```{r echo=T}
library(ggcorrplot)
corr <- a2housing |> dplyr::select(where(is.numeric)) |> cor(use = "pairwise.complete.obs")
ggcorrplot(corr)
```

## Spatial data
- Each row of `a2housing` has **location**, given by the address.
- For example, the first row is a house at "1015 S Main St, Ann Arbor, MI 48104".
- Suppose we want to place these houses on a map...?

## Geocoding
- Geocoding is the process of converting addresses into geographic coordinates (latitude and longitude).
- I geocoded these data for you using the US Census Geocoding API.
- The `lat` and `lon` columns contain the latitude and longitude of each house.

## Mapping the data
```{webr}
ggplot(a2housing) + geom_point(aes(long, lat), alpha = 0.1)
```

## Problems with this plot
- The points are not projected onto a map.
- Not easy to see patterns or even localize the points to places in Ann Arbor.

## `sf` package
- To work with spatial data in R, we can use the `sf` package.
```{webr}
library(sf)
```
- The `sf` package provides functions for reading, writing, and manipulating spatial data.

## Convert to `sf` object
- We can convert `a2housing` to an `sf` object using the `st_as_sf()` function.
- We need to specify the coordinates and the coordinate reference system (CRS).

```{r}
a2housing_sf <- a2housing |> filter(!is.na(long), !is.na(lat)) |>
  st_as_sf(coords = c("long", "lat"), crs = 4326)
```

## Coordinate reference system
- The CRS specifies how the coordinates relate to locations on the earth.
- The most common CRS is WGS 84, which is used by GPS and web mapping services.
- The code for WGS 84 is 4326.
- We will talk more about CRS next week.

## Plotting `sf` objects
We can plot `sf` objects using `geom_sf()`.
```{webr}
ggplot(a2housing_sf) + geom_sf(alpha = 0.1)
```

## Providing context
```{r}
library(ggspatial)
filter(a2housing_sf, sale_price < 2e6) |>
  ggplot() +  annotation_map_tile(type = "osm", zoom = 13) +
  geom_sf(aes(color=sale_price), alpha = 0.1) +  scale_color_viridis_c()

```
  
# Spatial autocorrelation
  > Everything is related to everything else, but near things are more related than distant things.
  - [Tobler's First Law of Geography](https://en.wikipedia.org/wiki/Tobler%27s_first_law_of_geography):

##
- Spatial autocorrelation is the correlation of a variable with itself through space.
- In other words, it is the tendency for nearby locations to have similar values.

## Visualizing spatial autocorrelation
```{r}
ggplot(a2housing_sf) + geom_sf(aes(color = log10(sale_price)), alpha = 0.5) + scale_color_viridis_c()
```

## Quantifying spatial autocorrelation
- We can see that there is spatial autocorrelation in the housing prices.
- "How correlated is my house price with my neighbor's?"
- To quantify this we need to locate the nearest neighbors of each house.

## Nearest neighbors
- We can use the `st_nearest_feature()` function from the `sf` package to find the
  nearest neighbor (in the dataset) of each house.
```{r eval=F}
a2housing_sf$nn <- st_nearest_feature(a2housing_sf)  # slow (why?)
```

## Nearest neighbors
- The `nn` column contains the index of the nearest neighbor for each house.
- For example, the nearest neighbor of the first house (index 1):
```{webr}
a2housing_sf$address[1]
a2housing_sf$address[a2housing_sf$nn[1]]
```

## Correlation with nearest neighbor
- We can compute the correlation between the sale price of each house and its nearest neighbor.
```{webr}
cor(a2housing_sf$sale_price, a2housing_sf$sale_price[a2housing_sf$nn], use = "complete.obs")
```

# Choropleth 
- A choropleth is a map that uses colors to represent values in different regions.
- We can create a choropleth of average housing prices by neighborhood.
- To do this we need to define "neighborhoods"

## Census tracts
- The US Census Bureau divides the country into small geographic areas called census tracts.
- Census tracts are designed to be relatively homogeneous in terms of population characteristics, economic status,
  and living conditions.

## Obtaining census tract data
- Use the excellent `tidycensus` package to obtain census tract shapefiles.
- Can also specific census variables (e.g. median hh income, `B19013_001`).
```{r}
library(tidycensus)
a2tracts <- get_acs(geography = "tract", variables = "B19013_001", state = "MI", 
                    county = "Washtenaw", geometry = TRUE)
```

## Plotting census tracts
```{webr}
ggplot(a2tracts) + geom_sf(aes(fill = estimate)) + scale_fill_viridis_c()
```
## Joining housing data to census tracts
- If we want to summarize housing prices by census tract, we need to join the two datasets.
- This is a new and special kind of join that we have not seen before: a spatial join.

## Idea of spatial join
```{r}
ggplot() + 
  geom_sf(data = a2housing_sf, aes(color = sale_price), alpha = 0.1) + 
  geom_sf(data = a2tracts, fill = NA) +
  scale_color_viridis_c() + 
  xlim(c(-83.80, -83.65)) + ylim(c(42.22, 42.32))
```

## Spatial join
- We can use the `st_join()` function from the `sf` package to perform a
  spatial join.
- This will join the attributes of the census tracts to the housing data based on location. 
```{webr}
st_join(a2housing_sf, a2tracts, join = st_within)
```

## 
- Why did we get this error?
- The two datasets have different coordinate reference systems (CRS).
```{webr}
st_crs(a2housing_sf)
st_crs(a2tracts)
```

## Transforming CRS
- We can use the `st_transform()` function to transform the CRS of one of the datasets
- Let's transform `a2tracts` to the CRS of `a2housing_sf`
```{r echo=T}
(a2ht <- a2tracts |> st_transform(st_crs(a2housing_sf)) |> 
   st_join(a2housing_sf, join = st_contains))
```

##

```{webr}
a2ht |> group_by(geometry) |> 
  summarize(mean_price = mean(sale_price, na.rm = TRUE)) |>
  ggplot() + geom_sf(aes(fill = mean_price)) + scale_fill_viridis_c()
```

## Learning more about `sf`
`sf` has a good [cheat sheet](https://rstudio.github.io/cheatsheets/sf.pdf) which
tells more about the neat things it can do.

