---
title: "DATASCI 306"
subtitle: "Lecture 10"
format: live-revealjs
engine: knitr
revealjs-plugins:
  - drop
resources:
  - data/a2weather.RData
webr:
  packages:
    - tidyverse
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(modelr)
load("data/a2weather.RData")
theme_set(theme_classic())
```
```{webr}
#| echo: false
#| message: false
#| warning: false
#| autorun: true
library(tidyverse)
library(modelr)
theme_set(theme_classic())
load("data/a2weather.RData")
```

[https://ds306.org/lectures/week5/lecture10.html](https://ds306.org/lectures/week5/lecture10.html)

## Recap from Lecture 9
- Advanced plots.
- Learn how to import data from the web.
- Study the labor market for college grads and data scientists.

## Next three lectures
- Study global climate trends.
- Linear regression.

## GHCN
- The [Global Historical Climate Network](https://www.ncei.noaa.gov/products/land-based-station/global-historical-climatology-network-daily) is amazing database of daily weather data going back, in some cases, >100 years. 
- I downloaded daily observation data for Ann Arbor from 1891 to present:
```{webr}
load("data/a2weather.RData")
```

## Exploring the data
- What do the columns mean?
- Are these data in wide format or long format?
- What are the units?
- How is missing data encoded?
- Are there any outliers/duplicates?

## Exploring the data
```{webr}
glimpse(a2weather)
```
- There are some standard columns: `id`, `date`, `year`, `month`
- What does `element` represent? What about the remaining columns?

## Tidying the data
A tidy version of this data frame will look like:

       id           year month element   day VALUE MFLAG QFLAG SFLAG
       <chr>       <int> <int> <chr>   <int> <dbl> <chr> <chr> <chr>
     1 USC00200230  1891    10 TMAX        1   206 " "   " "   6    
     2 USC00200230  1891    10 TMAX        2   267 " "   " "   6    

## Reshaping the data
- The data are in wide format, let's convert to long format.
- ChatGPT gives:
```{webr}
a2weather %>%
  pivot_longer(
    cols = starts_with("VALUE"),
    names_to = "day",
    names_prefix = "VALUE",
    values_to = "value"
  )
```

## Reshaping the data (improved version)
- ChatGPT did not do a good job -- not clear what all the columns now represent.
- With more careful prompting, we get:
```{webr}
a2weather %>%
  pivot_longer(
    cols = -(1:4),
    names_to = c(".value", "day"),
    # pattern: capture VALUE, MFLAG, QFLAG, SFLAG and the day number
    names_pattern = "(VALUE|MFLAG|QFLAG|SFLAG)(\\d+)",
    names_transform = list(day = as.integer)
  )
```

## Reshaping the data (improved version)
- ChatGPT did not do a good job -- not clear what all the columns now represent.
- With more careful prompting, we get:
```{webr}
a2w1 <- a2weather %>%
  pivot_longer(
    cols = -(1:4),
    names_to = c(".value", "day"),
    # pattern: capture VALUE, MFLAG, QFLAG, SFLAG and the day number
    names_pattern = "(VALUE|MFLAG|QFLAG|SFLAG)(\\d+)",
    names_transform = list(day = as.integer)
  ) |> print()
```

## Additional cleanups
- Convert dates
- Units?
- Missing data?
```{webr}
a2weather_tidy
```

## Preliminary exploration
How many observations are there per year?
```{webr}
a2weather_tidy |>
  mutate(year = year(date)) |>
  count(year) |>
  ggplot(aes(year, n)) +
  geom_line() +
  labs(
    title = "Number of Daily Observations per Year",
    x = "Year",
    y = "Number of Observations"
  )
```

## Daily temperature
```{webr}
a2weather_tidy |>
  ggplot(aes(x = date, y = tmax)) +
  geom_line() +
  labs(
    title = "Daily Maximum Temperature in Ann Arbor",
    x = "Date",
    y = "Temperature (°C)",
    color = "Element"
  )
```

## Monthly average temperature
```{webr}
a2weather_tidy |>
  mutate(month = floor_date(date, "month")) |>
  group_by(month) |>
  summarize(tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x = month, y = tmax)) +
  geom_line() +
  labs(
    title = "Monthly Average Maximum Temperature in Ann Arbor",
    x = "Month",
    y = "Temperature (°C)",
  )
```

## Recent temperature
```{webr}
a2weather_tidy |>
  filter(date >= today() - ddays(60)) |>
  ggplot(aes(x = date, y = tmax)) +
  geom_line() +
  labs(
    title = "Daily High Temperature in Ann Arbor",
    x = "Day",
    y = "Temperature (°C)",
  )
```
## Temperature on Halloween
```{webr}
a2weather_tidy |>
  filter(month(date) == 10, day(date) == 31) |>
  ggplot(aes(x = year(date), y = tmax * 1.8 + 32)) +
  geom_line() +
  labs(
    title = "Halloween High Temperature in Ann Arbor",
    x = "Year",
    y = "Temperature (°F)",
  )
```
## Are temperatures increasing?
- Now suppose we want to go further and quantify whether temperatures are increasing over time.
- To do this we need a *statistical model*.

# Statistical modeling

A statistical model is a mathematical formula that relates an outcome with one or more explanatory variables.

$$ Y = f (X) + 	\epsilon$$

* Y - outcome
* f - model function
* X - explainer
* $\epsilon$ - noise

## Linear regression
- The simplest and most widely used used statistical model is *linear regression*.
$$ y = \beta_0 + \beta_1 \times x + \epsilon$$
- i.e., $f(x) = \beta_0 + \beta_1(x)$ (a line)

## Linear regression versus other models
- Linear regression is easy to fit and interpret: $\beta_1$ is the change in $y$ for a one unit increase in $x$.
- In more advanced courses we will learn about more flexible models (e.g., splines, GAMs, random forests, neural networks).
- There is no "true" model, just models that are more or less useful for a given purpose.

## Fitting a linear regression model
- Let's use linear regression to quantify whether temperatures are increasing over time in Ann Arbor.
- Concretely, let's fit the following model:
    $$\mathrm{tmax}(t) = \beta_0 + \beta_1 \times t + \epsilon$$
    
## Linear regression in R
```{webr}
mdl <- lm(tmax ~ date, data = a2weather_tidy)
summary(mdl)
```

## Important components of the regression output
- Coefficients table
    - Estimate: the fitted values of $\beta_0$ and $\beta_1$
    - Std. Error: standard error of the estimates
    - Pr(>|t|): p-value for null hypothesis that coefficient is zero
- Multiple R-squared: proportion of variance explained by the model
- F-statistic: overall significance of the model
- Degrees of freedom: number of observations minus number of parameters

## Interpreting the results
```{webr}
coef(mdl)
```
- The estimated slope is 4.3e-5, which means that the daily maximum temperature increases by 0.000043 degrees Celsius per day.
- How much increase over 100 years?

## $R^2$
- R-squared measures the proportion of variance explained by the model.
- In simple linear regression, it is the square of the correlation between $x$ and $y$.
```{webr}
summary(mdl)$r.squared
cor(as.numeric(a2weather_tidy$date), a2weather_tidy$tmax, use="complete.obs")^2
```
- Important: $R^2$ does not measure whether the model is "true" or "useful". I can always
make $R^2$ larger by adding more variables.

## Residuals
- The residual is the difference between the predicted and observed value:
$$\text{resid}_i = y_i - \hat{y}_i = y_i - (\beta_0 + \beta_1 x_i).$$
- Thus, it's the part of the data that is not explained by the model.
- The residuals should look like random noise if the model is a good fit. 
- If there are clear patterns in the residuals, then the model is missing something.

## Residual plot
```{webr}
a2weather_tidy |>
  add_residuals(mdl) |>
  ggplot(aes(x = resid)) +
  geom_histogram() +
  labs(
    title = "Residuals from Linear Regression Model",
    x = "Residual",
  )
```

## Improving the model
- What is driving the patterns in the residuals? 
- Seasonal variation in temperature.
- Let's make the model slightly more advanced by adding in indicator variable for
  the summer:

$$
\mathrm{tmax}(t) = \beta_0 + \beta_1 \times t + \beta_2 \times I(\text{summer}) + \epsilon
$$
## Regression with categorical variables
- If we add a categorical variable with $k$ levels to a regression model, R will automatically create $k-1$ indicator variables.
- The coefficients of these indicator variables represent the difference in intercept between the corresponding level and the reference level.
```{webr}
lm(Sepal.Length ~ Species, data = iris) |> summary()
```


## Summer regression
- For each row we generate a 0/1 indicator variable that is 1 if the month is May to September (inclusive) and 0 otherwise.
- For the summer months, the model is $\mathrm{tmax}(t) = \beta_0 + \beta_1 \times t + \beta_2 + \epsilon$. The intercept is now $\beta_0 + \beta_2$.
- For the other months, the model is $\mathrm{tmax}(t) = \beta_0 + \beta_1 \times t + \epsilon$. The intercept is $\beta_0$.
  
## Graphically
In graphical terms, we're fitting two lines with the same slope but different intercepts.
```{r}
#| echo: false
#| message: false
#| warning: false
a2weather_tidy |>
  mutate(summer = ifelse(month(date) %in% 5:9, "Summer", "Other")) |>
  ggplot(aes(x = date, y = tmax, color = summer)) +
  geom_point(alpha = 0.01) +
  # make regression lines have same slope
  geom_smooth(method = "lm") +
  labs(
    title = "Daily Maximum Temperature in Ann Arbor",
    x = "Date",
    y = "Temperature (°C)",
    color = "Season"
  )
```
  
## Summer regression
```{webr}
mdl.summer <- lm(tmax ~ date + I(month(date) %in% 5:9), data = a2weather_tidy)
summary(mdl.summer)
```
Whoah! Our R^2 jumped to around 0.6!

## Interpreting the model
- When we add indicator variables, the interpretation of the coefficients changes.
- The slope $\beta_1$ is now the increase in temperature over time *after accounting for seasonal variation*.
- The coefficient $\beta_2$ is the increase in temperature during the summer months (May to September) compared to the other months, after accounting for the trend over time.
- The slope slightly lower than before, but still positive and significant.

## Residual plot
```{webr}
a2weather_tidy |>
  add_residuals(mdl.summer) |>
  ggplot(aes(x = resid)) +
  geom_histogram() +
  labs(
    title = "Residuals from Linear Regression Model with Summer Indicator",
    x = "Residual",
  )
```

## Seasonal trends
We can still do better -- let's look at the annual temperature cycle:
```{webr}
#| autorun: true
(p1 <- a2weather_tidy |> mutate(doy = yday(date)) |> group_by(doy) |>
  summarize(tmax = mean(tmax, na.rm = TRUE)) |>
  ggplot(aes(x = doy, y = tmax)) + geom_line() +
  labs(title = "Average Annual Temperature Cycle in Ann Arbor",
    x = "Day of Year", y = "Temperature (°C)"))
```

## Describing the seasonality
Let's try guessing this function "by hand":
```{webr}
p1 + stat_function(
  fun =  function(x) sin(x),
  color = "red")
```

## Adding nonlinear terms
- Learn the function from data
$$ \mathrm{tmax}(t) = \beta_0 + \beta_1 \times \sin(t / 365+\beta_2) + \epsilon $$
- Note that we have to estimate $\beta_2$ as well, which makes this a nonlinear regression model.
- *But*, we can be smart and use triginometry:
  $$A \sin(\omega t + \phi) = \beta_1 \sin(\omega t) + \beta_2 \cos(\omega t)$$
  where $A = \sqrt{\beta_1^2 + \beta_2^2}$ and $\phi = \tan^{-1}(\beta_2/\beta_1)$.

## Sine/cosine regression
```{webr}
mdl.trig <- a2weather_tidy |> mutate(doy = yday(date)) |>
  lm(tmax ~ sin(2 * pi * doy / 365) + cos(2 * pi * doy / 365), data = _)
summary(mdl.trig)
```

## Verify that it worked
```{webr}
beta <- coef(mdl.trig)
p1 + stat_function(fun = function(x)
  beta[1] + 
  beta[2] * sin(2 * pi * x / 365) + 
  beta[3] * cos(2 * pi * x / 365), color = "red")
```
## Putting it all together
```{webr}
mdl.final <- lm(tmax ~ date + sin(2 * pi * yday(date) / 365) + cos(2 * pi * yday(date) / 365), data = a2weather_tidy)
summary(mdl.final)
```

## 

```{r}
#| echo: false
set.seed(42)
mpg |> sample_frac(.9) |> 
  mutate(x1=cty, y=hwy, x2=fct_recode(
  class, 
    "A"="2seater",
    "A"="subcompact",
    "A"="compact",
    "B"="midsize",
    "C"="suv",
    "C"="minivan",
    "D"="pickup",
)) |> select(y, x1, x2) %>% lm(y ~ x1 + x2, data=.) |> summary()
```