---
title: "DATASCI 306"
subtitle: "Lecture 3"
format: live-revealjs
engine: knitr
webr:
  packages:
    - tidyverse
    - dslabs
  resources:
    - survey1.csv
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

## Loading this notebook

```{=html}
<style>
.qrcode img {
  margin: 20px auto;
}
</style>
```

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(dslabs)
```
```{webr}
#| echo: false
#| message: false
#| warning: false
#| autorun: true
library(tidyverse)
library(dslabs)
set.seed(1)
survey1 <- read_csv("survey1.csv")
```

[https://ds306.org/lectures/week2/lecture3.html](https://ds306.org/lectures/week2/lecture3.html)

## Recap from Lecture 2

Last lecture we covered:

-  Basic dpylr operations: `filter()`, `arrange()`, `select()`, `mutate()`
-  The pipe operator: `%>%` or `|>`


## Common mistake on quiz
`mean(df)` vs. `mean(df$column_name)`
```{webr}
mean(diamonds)
```


## Lecture 3

Today's lecture will cover:

-  Writing a function
-  More practice with `dplyr`
-  Working with tabular data
-  Hypothesis testing


## Loading data into R

```{webr}
# download.file("https://websites.umich.edu/~jonth/stats306/datasets/survey1.csv", "survey1.csv")
survey1 <- read_csv("survey1.csv")
```

## Cleaning up the height data
From ChatGPT:
```{r}
#| eval: false
#| echo: true
height_cm <- case_when(
  raw < 1 ~ raw * 100000,                    # kilometers → cm
  raw >= 1.7 & raw < 2.1 ~ raw * 100,        # meters → cm
  raw > 100 & raw < 250 ~ raw,               # centimeters (already cm)
  raw >= 48 & raw <= 90 ~ raw * 2.54,        # inches → cm
  raw > 400 & raw < 800 ~ {
    feet <- floor(raw / 100)
    inches <- raw %% 100
    (feet * 12 + inches) * 2.54              # feet+inches concatenated → cm
  },
  raw >= 4 & raw <= 8 ~ {
    feet <- floor(raw)
    inches <- (raw - feet) * 100
    (feet * 12 + inches) * 2.54              # feet.inches decimal → cm
  },
  TRUE ~ NA
)
```

## Converting "decimal feet"
-  Some people reported their heights in the format "5.8" (5 feet 8 inches), or 
"5.11" (5 feet 11 inches), or "6.0" (6 feet 0 inches).
-  ChatGPT suggested the following code to convert these to cm:
```{r}
#| eval: false
#| echo: true
feet <- floor(height)
inches <- (height - feet) * 100
(feet * 12 + inches) * 2.54              # feet.inches decimal → cm
```

## Functions in R
-  To define a function in R, we use the syntax:
```{r eval=F, echo=T}
my_function <- function(arg1, arg2, ...) {
  # code that does something with arg1, arg2, ...
  return(some_value)  # return() is optional
}
```
-  The function can then be called with:
```{r eval=F, echo=T}
my_function(value1, value2, ...)
```
## R functions vs. other languages
-  `return()` is optional. R will return the last evaluated expression.
-  Functions are just variables: `function_name <- function(...) { ... }`
-  R functions can return multiple values using a vector or list.

## Function to convert "decimal feet" to cm
- Let's try running this code with a few different values of `height`.
- To conveniently do this, we can define a _function_:
```{webr}
unpack_dft <- function(height) {
  # "decimal feet" -> cm
  feet <- floor(height)
  inches <- (height - feet) * 100
  c(feet, inches)  # return both feet and inches
}
```

## Calling a function
- Now we can easily test this function with different values of `height`:
```{webr}
unpack_dft(5.8)
unpack_dft(5.11)
unpack_dft(6.0)
```

## Function to convert "decimal feet" to cm
```{webr}
dft_to_cm <- function(height) {
  parts <- unpack_dft(height)
  feet <- parts[1]
  inches <- parts[2]
  (feet * 12 + inches) * 2.54              # feet.inches decimal → cm
}
```

## Using our function in `mutate()`
```{webr}
raw_height_to_cm <- function(raw) {
  case_when(
    raw < 1 ~ raw * 100000,                   # kilometers → cm
    raw >= 1.7 & raw < 2.1 ~ raw * 100,       # meters → cm
    raw > 100 & raw < 250 ~ raw,              # centimeters (already cm)
    raw >= 48 & raw <= 90 ~ raw * 2.54,       # inches → cm
    raw > 400 & raw < 800 ~ dft_to_cm(raw / 100),   # feet+inches → cm
    raw >= 4 & raw <= 8 ~ dft_to_cm(raw),           # feet.inches decimal → cm
    TRUE ~ NA_real_
  )
```

## Rewriting our pipeline
```{webr}
survey1_final <- survey1 |>
  rename(timestamp=1, sex=2, height=3) |> 
  mutate(height_cm = raw_height_to_cm(height)) |> 
  arrange(height_cm) |> 
  select(sex, height_cm)
glimpse(survey1_final)
```

# Hypothesis testing

## Testing for difference in means
-  In STATS 250 you learned how to test hypotheses about population means.
-  We will now learn how to do this in R.
-  We will use the `t.test()` function to test hypotheses about means.

## Example data
```{webr}
x1 <- rnorm(100, mean=0, sd=1)
x2 <- rnorm(100, mean=1, sd=1)
```
- Are the means of `x1` and `x2` different?

## `t.test()`
```{webr}
t.test(x1, x2)
t.test(x1, x2, var.equal=TRUE)  # assume equal variances
```
**No need** to manually calculate the t-statistic or null distribution -- you've graduated past that.

## Example: testing average height
Are the male and female heights in the class equal?
```{webr}
male.heights <- NA
female.heights <- NA
# perform test
```

## 💡 Connection to regression
- You also learned how to estimate a linear regression model in STATS 250.
- A t-test is equivalent to a regression with a binary predictor:

$$\texttt{height} = \alpha + \beta \times \texttt{is_male} + \epsilon$$

```{webr}
lm(height_cm ~ sex, survey1_final) |> summary()
```

## Testing for differences in proportions
-  We can also use R to test hypotheses about proportions.
-  We will use the `prop.test()` function to test hypotheses about proportions.

## Example data
Simulating a coin toss:
```{webr}
x <- rbinom(100, 1, 0.5)  # 100 tosses of a fair coin
```

## Testing a table
- `prop.test()` will also take a table as input:
```{webr}
prop.test(table(x))
```
- This is easier/less error prone.

## Gender ratio in this class?
- What is the ratio of male:female students in this class?
$$
H_0: \text{Ratio} = 1:1 \\
H_a: \text{Ratio} \neq 1:1
$$
```{webr}
# perform test for sex
```

## A second data set

<https://ds306.org/survey2>

{{< qrcode https://ds306.org/survey2 width=300 height=300 >}}

## Bivariate tests
- We can also test for independence between two categorical variables.
```{webr}
# simulate data
x <- sample(c("A", "B"), 100, replace=TRUE)
y <- sample(c("C", "D"), 100, replace=TRUE)
tab <- table(x, y)
tab
```

## Testing for independence
```{webr}
chisq.test(tab)
```

## Load survey 2
```{webr}
download.file("https://websites.umich.edu/~jonth/stats306/datasets/survey2.csv")
survey2 <- read_csv("survey2.csv")
```

## Testing for independence
```{webr}
# to be filled in in lecture
```