---
title: "DATASCI 306"
subtitle: "Lecture 15"
format: live-revealjs
engine: knitr
resources:
  - data/midwest.RData
  - data/temps.rds
  - data/poi.RData
  - data/dp.RData
  - data/a2landmarks.RData
  - data/a2parks.RData
  - data/vert.RData
  - data/min_park_dist.RData
  - data/a2grocery.RData
webr:
  packages:
    - tidyverse
    - sf
    - geosphere
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(tidyterra)
library(sf)
library(dodgr)
library(ggspatial)
load("data/midwest.RData")
load("data/poi.RData")
load("data/a2landmarks.RData")
load("data/dp.RData")
load("data/a2parks.RData")
load("data/a2graph.RData")
load("data/a2grocery.RData")
theme_set(theme_classic())
options(tigris_use_cache = TRUE)
```
```{webr}
#| echo: false
#| message: false
#| warning: false
#| autorun: true
library(tidyverse)
library(sf)
library(terra)
library(tidyterra)
load("data/midwest.RData")
load("data/poi.RData")
load("data/a2landmarks.RData")
load("data/dp.RData")
load("data/a2parks.RData")
load("data/vert.RData")
load("data/min_park_dist.RData")
load("data/a2grocery.RData")
theme_set(theme_classic())
sf_use_s2(FALSE)
```

```{=html}
<style>
.qrcode img {
  margin: 20px auto;
}
</style>
```

[https://ds306.org/lectures/week9/lecture15.html](https://ds306.org/lectures/week9/lecture15.html)


# Walk scoring

- Goal: Calculate walk scores based on distance to points of interest (POI) such as grocery stores, parks, schools, etc.
- Data: Locations of POIs with latitude and longitude coordinates.

## Measuring distances
Last lecture we learned how to measure distance in two ways:

- Euclidean distance: $\sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$
- Great-circle distance: shortest path along Earth's surface.

## Measuring distances

```{webr}
a2landmarks <- tribble(
  ~name, ~lat, ~long,
  "CHEM1400", 42.27847635024672, -83.73700902678496,
  "The Big House", 42.26601908368355, -83.7487063315362,
  "Diag", 42.27708990934726, -83.73817826048145,
  "North Campus", 42.29666404150472, -83.71033250016994,
  "The Arb", 42.28105377389016, -83.72566111988131
  )
a2landmarks_sf <- a2landmarks |> st_as_sf(coords = c("long", "lat"), crs = 4326)
```

## 

```{r}
#| echo: true
#| messages: false
a2landmarks_sf <- a2landmarks |> st_as_sf(coords = c("long", "lat"), crs = 4326)
a2landmarks_sf |> st_as_sf(coords = c("long", "lat"), crs = 4326) |>
  ggplot() + annotation_map_tile(zoom=14) + geom_sf(color="red") +
  # add place labels 
  geom_sf_text(aes(label=name), nudge_y = 0.001)
```
  
## Measuring distances

To measure the great circle distance, we use `st_distance`:

```{r}
#| echo: true
a2landmarks_sf |> st_distance() 
```

## Drawing great circle arcs

```{webr}
# use gcIntermediate then convert to an SF 
library(geosphere)
arcs <- geosphere::gcIntermediate(
  p1 = cbind(a2landmarks$long, a2landmarks$lat)[1, ],
  p2 = cbind(a2landmarks$long, a2landmarks$lat)[2, ],
  n = 50, addStartEnd = TRUE, sp = TRUE
) |> st_as_sf()
```


##

```{webr}
#| warning: false
ggplot() + geom_sf(data=arcs, color="blue", size=1) +
  geom_sf_text(data=a2landmarks_sf, aes(label=name), nudge_y = 0.001)
```

##

```{r}
#! echo: false
#| messages: false
#| warning: false
ggplot() + 
  annotation_map_tile(zoom=14) +
  geom_sf(data=arcs, color="blue", size=1) +
  geom_sf_text(data=a2landmarks_sf, aes(label=name), nudge_y = 0.001) 
```

## Differences at large scale

```{webr}
# plot gc dist between ann arbor and the eiffel tower
library(ggspatial)
arcs2 <- geosphere::gcIntermediate(
  # ann arbor
  p1 = cbind(a2landmarks$long, a2landmarks$lat)[1, ],
  # eiffel tower
  p2 = c(2.2945, 48.8584),
  n = 500, addStartEnd = TRUE, sp = TRUE
) |> st_as_sf()
ggplot() + geom_sf(data=arcs2, color="blue", size=1) 
```
## 

```{r}
#| echo: false
#| messages: false
#| warning: false
# plot gc dist between ann arbor and the eiffel tower
library(ggspatial)
arcs2 <- geosphere::gcIntermediate(
  # ann arbor
  p1 = cbind(a2landmarks$long, a2landmarks$lat)[1, ],
  # eiffel tower
  p2 = c(2.2945, 48.8584),
  n = 500, addStartEnd = TRUE, sp = TRUE
) |> st_as_sf()
ggplot() + geom_sf(data=arcs2, color="blue", size=1) 
```

# Walking distance

```{r}
#| echo: false
#| messages: false
#| warning: false
#| eval: true
# use gcIntermediate then convert to an SF 
geosphere::gcIntermediate(
  p1 = cbind(a2landmarks$long, a2landmarks$lat)[1, ],  # CHEM1400
  p2 = cbind(a2landmarks$long, a2landmarks$lat)[4, ],  # North Campus
  n = 50, addStartEnd = TRUE, sp = TRUE
) |> st_as_sf() |> ggplot() + 
  annotation_map_tile(zoom=14) +
  geom_sf(color="blue", size=1) +
  geom_sf_text(data=a2landmarks_sf, aes(label=name), nudge_y = 0.001)
```


# Walking distance

To measure walking distance we need:
- A dataset of streets, sidewalks, and paths;
- A way to calculate shortest paths along the street network

## Street network data
- [OpenStreetMap (OSM)](https://www.openstreetmap.org/#map=16/42.28098/-83.74897) is a free, open-source map of the world created by volunteers.
- We can download street network data from OSM using the `osmdata` R package or other tools.

##

```{webr}
ggplot(a2walk) + geom_sf(color="brown")
```

## 

```{r}
ggplot(a2walk) + annotation_map_tile(zoom=13) + geom_sf(color="brown")
```

## Walking from the diag to North Campus
- Using a [mapping application](https://www.openstreetmap.org/directions?engine=fossgis_osrm_car&route=42.2771027%2C-83.7380139%3B42.2942142%2C-83.7100389#map=15/42.28460/-83.72394) route from the Diag to North Campus is about 3.l km and takes about 39 minutes to walk.
- How to find the best route?

##

```{r}
a2landmarks_sf |> filter(name %in% c("Diag", "North Campus")) |>
ggplot() + geom_sf(data=st_crop(a2walk, st_bbox(a2landmarks_sf))) + geom_sf(color="red", size=5) +
  geom_sf_text(aes(label=name), nudge_y = 0.001)
```

## Routing algorithm
- We'll use `dodgr` package to calculate shortest paths along the street network.
- CS nerd stuff:
  - Technically this is solving a graph theory problem called ["shortest path"](https://en.wikipedia.org/wiki/Shortest_path_problem).
  - Dijkstra's algorithm is a classic algorithm for solving shortest path problems on graphs with non-negative edge weights.
  
  
##  Finding a shortest path using `dodgr`

```{r}
#| eval: false
#| echo: true
# diag to north campus
diag = a2landmarks[3, c("lat", "long")]
north_campus = a2landmarks[4, c("lat", "long")]
# solve the routing problem; result is a list of vertices
dp <- dodgr_paths(a2graph_w, from = diag, to = north_campus)
verts <- dodgr_vertices(a2graph_w)
path1 <- verts [match (dp [[1]] [[1]], verts$id), ]
```

##

```{r}
st_as_sf(path1, coords = c("x", "y"), crs = 4326) |> 
  ggplot() +  annotation_map_tile(zoom=14) + geom_sf(color="blue", size=1)
```

## Creating a line
```{webr}
st_as_sf(path1, coords = c("x", "y"), crs = 4326) |> st_combine() |> 
  st_cast("LINESTRING") |>
  ggplot() + geom_sf(color="blue", size=1)
```

## 
```{r}
#| echo: false
st_as_sf(path1, coords = c("x", "y"), crs = 4326) |> st_combine() |> 
  st_cast("LINESTRING") |>
  ggplot() + annotation_map_tile(zoom=14) +
  geom_sf(color="blue", size=1)
```

# Measuring walking distance to amenities
Now that we know how to measure walking distance, we can calculate walk scores based on distance to points of interest (POI) such as grocery stores, parks, schools, etc.

## Parks in Ann Arbor
Ann Arbor has >160 parks, helpfully mapped by the city government and released to the public.

```{webr}
ggplot(a2parks) + geom_sf(data=a2walk) + geom_sf(fill="green")
```


## Walking distance from a point to a park
```{r}
#| echo: true
#| warning: false
# pick a point in Ann Arbor
chem1400 <- a2landmarks |> filter(name == "CHEM1400") |> select(y=lat, x=long)
# pick a park
park <- a2parks |> filter(NAME == "Gallup Park") |> st_centroid() |> st_coordinates()
# calculate walking distance
vert <- dodgr_vertices(a2graph_w)
chem_to_park <- dodgr_paths(a2graph_w, from = park, to = chem1400)
```

## Is this correct?
```{r}
chem_park_path <- vert [match (chem_to_park [[1]] [[1]], vert$id), ]
st_as_sf(chem_park_path, coords = c("x", "y"), crs = 4326) |> st_combine() |> 
  st_cast("LINESTRING") |> ggplot() + annotation_map_tile() + geom_sf(color="blue", size=1)
```

## What happened?

```{webr}
a2parks
```

## Corrected walking distance
```{r}
# pick a point in Ann Arbor
chem1400 <- a2landmarks |> filter(name == "CHEM1400") |> select(y=lat, x=long)
# pick a park
park <- a2parks |> filter(NAME == "Gallup Park") |> st_transform(4326) |> st_centroid() |> st_coordinates()
# calculate walking distance
vert <- dodgr_vertices(a2graph_w)
chem_to_park <- dodgr_paths(a2graph_w, from = park, to = chem1400)
```

## Corrected plot

```{r}
#| echo: false
chem_park_path <- vert [match (chem_to_park [[1]] [[1]], vert$id), ]
st_as_sf(chem_park_path, coords = c("x", "y"), crs = 4326) |> st_combine() |> 
  st_cast("LINESTRING") |> ggplot() + annotation_map_tile() + geom_sf(color="blue", size=1)
```

## Plotting the distance to the nearest park

```{r}
#| message: false
#| warning: false
a2grid <- st_make_grid(st_transform(a2parks, 4326), cellsize=0.001) |> st_sf()
a2grid_coords <- a2grid |> st_centroid() |> st_coordinates()
park_coords <- a2parks |> st_transform(4326) |> st_centroid() |> st_coordinates()
a2grid_dist <- dodgr_distances(a2graph_w, from = a2grid_coords[, c("X", "Y")], to = park_coords)
min_park_dist <- apply(a2grid_dist, 1, function(x) min(x, na.rm=T))
```

##
```{webr}
a2grid |> mutate(min_park_dist = min_park_dist) |>
  ggplot() + geom_sf(aes(fill=min_park_dist), color=NA) + 
  scale_fill_viridis_c()
```

## Walkability to other amenities
Grocery stores:
```{r}
#| eval: false
#| echo: true
library(osmdata)
a2bb <- getbb("Ann Arbor, MI")
query_bbox <- opq(bbox = a2bb) %>%
  add_osm_feature(key = "shop", value = c("supermarket", "grocery"))
a2_grocery_stores <- osmdata_sf(query_bbox)
```

##

```{webr}
ggplot(a2grocery) +  geom_sf(a2walk, color="brown") + geom_sf(color="orange", size=3)
```

## Walk distance to nearest grocery store

```{r}
#| message: false
#| warning: false
#| eval: true
a2grocery_coords <- a2grocery |> st_transform(4326) |> st_coordinates()
a2grid_dist_grocery <- dodgr_distances(a2graph_w, from = a2grid_coords, to = a2grocery_coords)
min_grocery_dist <- apply(a2grid_dist_grocery, 1, function(x) min(x, na.rm=T))
```

##

```{r}
#| message: false
a2grid |> mutate(min_grocery_dist = min_grocery_dist) |>
  ggplot() +  annotation_map_tile(zoom=13) +
  geom_sf(aes(fill=min_grocery_dist), color=NA, alpha=.9)
```




## Quiz questions

1. How do walk scoring systems typically measure walkability?
   - A) By calculating Euclidean distances to amenities.
   - B) By calculating great-circle distances to amenities.
   - C) By calculating walking distances along street networks to amenities.
   - D) By counting the number of amenities within a fixed radius.
   
2. One reason why great-circle distance is not suitable for walk scoring is:
   - A) It does not account for elevation changes.
   - B) It assumes a flat Earth.
   - C) It does not consider the actual paths people can take on streets and sidewalks.
   - D) It is computationally intensive.
   
3. 