---
title: "DATASCI 306"
subtitle: "Lecture 12"
format: live-revealjs
engine: knitr
resources:
  - data/india-TAVG-Trend.txt
  - data/weather.RData
  - data/globalclimate.RData
webr:
  packages:
    - tidyverse
    - slider
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(modelr)
library(slider)
load("data/weather.RData")
load("data/globalclimate.RData")
theme_set(theme_classic())
```
```{webr}
#| echo: false
#| message: false
#| warning: false
#| autorun: true
library(tidyverse)
library(modelr)
library(slider)
theme_set(theme_classic())
load("data/weather.RData")
load("data/globalclimate.RData")
```

[https://ds306.org/lectures/week6/lecture12.html](https://ds306.org/lectures/week6/lecture12.html)

## Recap from Lecture 11 
- Regression modeling.
- Trends in the local weather of Ann Arbor and Reno.

## Today's lecture
- More on regression modeling.
- Trends in global temperatures.

## Regression with interaction terms
- Let's consider the model
$$ \mathrm{tmax} = \beta_0 + \beta_1 \times \mathrm{date} + \beta_2 \times \mathrm{city} + \epsilon $$
- Here, `city` is a categorical variable with two levels: "Ann Arbor" and "Reno".
- The coefficient $\beta_2$ measures the average difference in temperature between the two cities.

## 

```{webr}
mdl1 <- lm(tmax ~ date + city, data = combined)
summary(mdl1)
```
This tells us that Reno is on average about 4.8°C warmer than Ann Arbor, which is 
not surprising.

## Understanding the model

When we include a categorical vector like `city` in a regression model, R generates a new column in our data:

```{webr}
model_matrix(combined, tmax ~ date + city)
```

## Interpreting coefficients

- When `city` is "Ann Arbor", the model is
$$ \mathrm{tmax} = \beta_0 + \beta_1 \times \mathrm{date} + 0 + \epsilon  $$
- When `city` is "Reno", the model is
$$
\begin{align}
\mathrm{tmax} &= \beta_0 + \beta_1 \times \mathrm{date} + \color{red}{ \beta_2} + \epsilon\\
&= (\beta_0 + \color{red}{\beta_2}) + \beta_1 \times \mathrm{date} + \epsilon
\end{align}
$$
- So $\color{red} {\beta_2}$ is the difference in average temperature between Reno and Ann Arbor.

## Interpreting the model (graphically)
We fit two regression lines with the same slope:

```{r}
#| echo: false
#| warning: false
#| message: false
mdl <- lm(tmax ~ date + city, data = combined)

ggplot(combined, aes(date, tmax, color = city)) +
  geom_point(alpha = 0.01) +
  geom_abline(intercept = coef(mdl)[1], slope = coef(mdl)[2], color = "blue") +
  geom_abline(intercept = coef(mdl)[1] + coef(mdl)[3], slope = coef(mdl)[2], color = "red") +
  scale_color_manual(values = c("blue", "red")) +
  labs(y = "Daily maximum temperature (°C)", x = "Date")    
```




## Adding interaction terms
- Now consider the model
$$ \mathrm{tmax} = \beta_0 + \beta_1 \times \mathrm{date} + \beta_2 \times \mathrm{city} + \beta_3 \times \mathrm{date} \times \mathrm{city} + \epsilon  $$
- The new term $\beta_3 \times \mathrm{date} \times \mathrm{city}$ is an **interaction term**.
- The coefficient $\beta_3$ measures the difference in warming rates between the two cities.

## Interpreting the model (nuts & bolts version)
R generates two new columns in our data:
```{webr}
csub <- combined |> sample_n(5)
modelr::model_matrix(csub, tmax ~ date * city) |> head()
```

## Interpreting the model (math version)
- When `city` is "Ann Arbor", the model is
$$ \mathrm{tmax} = \beta_0 + \beta_1 \times \mathrm{date} + 0 + 0 + \epsilon  $$
- When `city` is "Reno", the model is
$$ \mathrm{tmax} = \beta_0 + \beta_1 \times \mathrm{date} + \color{red}{\beta_2} + \color{blue}{\beta_3 \times \mathrm{date}} + \epsilon  $$
which can also be written as
$$ \mathrm{tmax} = (\beta_0 + \color{red}{\beta_2}) + (\beta_1 + \color{blue}{\beta_3}) \times \mathrm{date} + \epsilon  $$

## Interpreting the model (picture version)
We fit two regression lines with different slopes:
```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(combined, aes(date, tmax, color = city)) +
  geom_point(alpha = 0.01) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("blue", "red")) +
  labs(y = "Daily maximum temperature (°C)", x = "Date")
```

# Surface temperatures
- Now let's consider global surface temperatures.
- We'll use the BEST dataset, which is available from [https://berkeleyearth.org/data/](https://berkeleyearth.org/data/).


## Data format
The [BEST data files](https://berkeley-earth-temperature.s3.us-west-1.amazonaws.com/Regional/TAVG/india-TAVG-Trend.txt) have the following format:

```
%                  Monthly          Annual          Five-year        Ten-year        Twenty-year
% Year, Month,  Anomaly, Unc.,   Anomaly, Unc.,   Anomaly, Unc.,   Anomaly, Unc.,   Anomaly, Unc.
 
  1816     5    -0.542  1.344       NaN    NaN       NaN    NaN       NaN    NaN       NaN    NaN
  1816     6    -1.214  1.506       NaN    NaN       NaN    NaN       NaN    NaN       NaN    NaN
  ...
```

## Fixed-width data
- This is an example of a fixed-width data file.
- Unlike CSV, the columns are not separated by commas.
- Instead, each column has a fixed width (number of characters).

## Reading fixed-width data
(Pre-2020) Laboriously count the number of characters in each column, then use `read.fwf()`.
```{webr}
read_fwf("data/india-TAVG-Trend.txt")
```

## Reading fixed-width data
(Modernity) ChatGPT can figure this out
```{webr}
read_fwf("data/india-TAVG-Trend.txt", 
 fwf_cols(year = c(1, 6), month = c(7, 14), 
          anomaly = c(17, 23), uncert = c(25, 30)), comment = "%")
```

## Full climate data set
```{webr}
anomaly
```

## Investigating the raw data
```{webr}
ggplot(anomaly, aes(date, anomaly, color=region)) + geom_line() +
  labs(x = "Year", y = "Temperature anomaly (°C)")
```

## Reducing noise
```{webr}
anomaly |>  group_by(region, year(date)) |>  
  summarize(anomaly = mean(anomaly, na.rm=TRUE)) |> 
  ggplot(aes(`year(date)`, anomaly, color=region)) + geom_line()
```

## Regression questions
- An average, how much is global temperature increasing each year?
- Is the simple linear model adequate?

## Two-period linear model
Let's allow the slope to be different before and after 1960:
```{r}
#| echo: false
#| warning: false
#| message: false
anomaly |> filter(region == "united-states") |>  mutate(year = year(date)) |> 
  ggplot() +  geom_line(aes(date, anomaly), alpha=0.1) + 
  geom_smooth(aes(date, anomaly), method="lm", formula = y ~ x * (x > 1960), se=T)
```

## 
```{webr}
anomaly |> filter(region == "united-states") |> 
  lm(anomaly ~ date * I(year(date) > 1960), data = _) |> summary()
```
## A three-way interaction
```{webr}
anomaly |> lm(anomaly ~ date * I(year(date) > 1960) * region, data = _) |> summary()
```

## Graphical interpretation
```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(anomaly, aes(date, anomaly, color=region)) + 
  geom_smooth(method="lm", formula = y ~ x * (x > 1960), se=F)
```

# Correlation between temperature and CO2 levels
Atmospheric CO2 over time: ([Mauna Loa data](https://scrippsco2.ucsd.edu/data/atmospheric_co2/primary_mlo_co2_record.html))
```{webr}
co2
```

## Atmospheric CO2 levels over time
```{webr}
ggplot(co2) + geom_line(aes(date, average)) +  labs(x = "Year", y = "CO2 (ppm)")
```

## Error bars
```{webr}
ggplot(co2) + geom_line(aes(date, average)) +  
  geom_ribbon(aes(date, ymin=average - sdev, ymax=average + sdev), alpha=0.2) +
  labs(x = "Year", y = "CO2 (ppm)")
```

## CO2 and temperature
```{webr}
co2_temp <- inner_join(anomaly |> filter(region == "global") |> select(date, anomaly), 
                       co2 |> select(date, average_co2), by = "date")
ggplot(co2_temp) + geom_point(aes(average_co2, anomaly), alpha=0.1) +
  labs(x = "CO2 (ppm)", y = "Temperature anomaly (°C)") + 
  geom_smooth(aes(average_co2, anomaly), method="lm", se=F)
```

## Regression model
```{webr}
mdl1 <- lm(anomaly ~ average_co2, data = co2_temp)
summary(mdl1)
```

## Change in CO2 over time
```{webr}
mdl2 <- lm(average_co2 ~ poly(date, 2), data = co2_temp)
summary(mdl2)
```

## Graphically
```{webr}
ggplot(co2_temp, aes(x=date, y=average_co2)) + 
  geom_line() + geom_smooth(method="lm", formula=y ~ poly(x, 2), se=F)
```

